---
title: "Collins_bingestats"
format: html
editor: visual
---

## Male vs Female Binge Stats Comparison

Stats for each figure appear first, followed by the graph

Note:

-   for behavioral intoxication time points, the first number indicates the day, the second the time point. For instance, beh_2_3 means day 2 time point 3 (11pm)

-   for withdrawal time points, the number indicates the hours after the last dose of ethanol. For instance, with_15 means 15 hours following the last dose of ethanol.

```{r}
library(tidyverse)
library(ggprism) #for prism theme
library(lmerTest) #for repeated measures anovas 
library(emmeans) #for post hoc test
library(rstatix) #for getsummarystats
```

```{r}
#Loading in data 

binge_stats <- read.csv("binge_master_file.csv")

#creating a unique_id column for the repeated measures anovas coming up 
binge_stats <- binge_stats %>%
  mutate(unique_id = paste(study, subject_id, sep = "_"))

#filtering out rats that died early, anything that's not an ethanol
binge_stats_filtered <- binge_stats %>%
filter(early_death == "no", treatment == "ethanol"
       ,study %in% c("srtx1", "srtx2", "srtx3", "srtx4", "big") #exclusing gub2
       #,binge_location == "kentucky "
       )

#counting males vs females by study
binge_stats_filtered %>%
  group_by(study) %>%
  summarize(
    female_count = sum(sex == "female"),
    male_count = sum(sex == "male")
  )

#counting males vs females by study
binge_stats_filtered %>%
  group_by(binge_location) %>%
  summarize(
    female_count = sum(sex == "female"),
    male_count = sum(sex == "male")
  )

#counting males vs females by study
binge_stats_filtered %>%
  summarize(
    female_count = sum(sex == "female"),
    male_count = sum(sex == "male")
  )

  
```

**Figure 1B: Behavioral Intoxication Over Time**

```{r}
#Creating Data Frame 
behav_intox <- binge_stats_filtered %>%
  select(6, 20:31, 50) %>% #selecting relevant columns of binge_stats
  pivot_longer(
    cols = -c(unique_id, sex),
    names_to = "time",
    values_to = "intox_score"
  ) 

#creating summary statistics 
behav_intox_grouped <- behav_intox %>%
  group_by(sex, time) %>%
  get_summary_stats(intox_score, show = c("mean", "sd", "se", "ci"))
print(behav_intox_grouped, n=100)

#creating model and running anova
model_behavintox <- lmer(intox_score ~ sex * time + (1 | unique_id), data = behav_intox)
model_behavintox_anova <- anova(model_behavintox)
print(model_behavintox_anova)

#posthoc test
emmeans_behavintox <- emmeans(model_behavintox, ~sex | time)
emmeans_behavintox_results <- pairs(emmeans_behavintox)
print(emmeans_behavintox_results)

#graph of behavioral intoxication over time 
ggplot(data = behav_intox_grouped, aes(x = time, y = mean, color = sex, group = sex)) + 
  geom_line(size = 1) + 
  geom_point() + 
  geom_errorbar(aes(y = mean, ymin = mean - se, ymax = mean + se, fill = sex), alpha = .5, width = 0.5) + 
  ylim(-.25, 5) + 
  scale_x_discrete(labels=c("beh_1_1" = "", 
                            "beh_1_2" = "Day 1",
                            "beh_1_3" = "",
                            "beh_2_1" = "", 
                            "beh_2_2" = "Day 2",
                            "beh_2_3" = "",
                            "beh_3_1" = "", 
                            "beh_3_2" = "Day 3",
                            "beh_3_3" = "",
                            "beh_4_1" = "", 
                            "beh_4_2" = "Day 4",
                            "beh_4_3" = ""
                            )) +
  ylab("Mean Behavioral Intoxication Score") + 
  xlab("Ethanol Treatment") + 
  theme_prism() +
  theme(legend.position = "none")

results_path <- "binge_stats_graphs"

ggsave(
  "/behav_intox_time.png",
  path = results_path,
  height = 5,
  width = 8.8
  )

```

**Figure 1C: Mean Behavioral Intoxication**

```{r}
#calculating summary statistics 
behav_intox_grouped_average<- behav_intox %>%
  group_by(sex, unique_id) %>%
  summarize(mean_intox_score = mean(intox_score))

#data frame for the jittered points
behav_intox_average <- behav_intox %>%
  group_by(unique_id, sex) %>%
  summarize(mean = mean(intox_score))

#calculating summmary statistics 
behav_intox_grouped_average_stats <- behav_intox_grouped_average %>%
  get_summary_stats(mean_intox_score, show = c("mean", "sd", "se", "ci"))
behav_intox_grouped_average_stats

#t.test
behav_intox_t.test <- t.test(mean_intox_score ~ sex, data = behav_intox_grouped_average)
behav_intox_t.test

#mean behavioral intoxication graph 
ggplot(data = behav_intox_grouped_average_stats, aes(x = sex, y = mean)) + 
  geom_bar(aes(fill = sex), stat = "summary", fun = "mean", alpha = 0.5, width = 0.5, color = "black") + 
  geom_jitter(data = behav_intox_average, aes(x=sex, y = mean, color = sex), alpha = 0.5, stroke = NA, width = 0.25) + 
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se),
                width = 0.2,
                color = "black",
                size = 0.5) +  
  theme_prism() + 
  ylab("Mean Behavioral Intoxication Score") +
  scale_x_discrete(labels=c("female" = "Female", 
                            "male" = "Male")) + 
  xlab("Sex") + 
  scale_y_continuous(limits = c(0,5), expand = c(0, 0)) + 
  theme(legend.position = "none")

ggsave(
  "/behav_intox_mean.png",
  path = results_path,
  height = 4,
  width = 2.5
  )

```

**Figure 1D: Peak Behavioral Intoxication**

```{r}
#getting the max intox score
behav_intox_grouped_peak<- behav_intox %>%
  group_by(sex, unique_id) %>%
  summarize(max_intox_score = max(intox_score)) #calculating max intox score 

#data frame for the jittered points
behav_intox_peak <- behav_intox %>%
  group_by(unique_id, sex) %>%
  summarize(mean = max(intox_score))
  
#calculating summmary statistics 
behav_intox_grouped_peak_stats <- behav_intox_grouped_peak %>%
  get_summary_stats(max_intox_score, show = c("mean", "sd", "se", "ci"))
behav_intox_grouped_peak_stats

#t.test
behav_intox_peak_t.test <- t.test(max_intox_score ~ sex, data = behav_intox_grouped_peak)
behav_intox_peak_t.test

#mean behavioral intoxication graph 
ggplot(data = behav_intox_grouped_peak_stats, aes(x = sex, y = mean)) + 
  geom_bar(aes(fill = sex), stat = "summary", fun = "mean", alpha = 0.5, width = 0.5, color = "black") + 
  geom_jitter(data = behav_intox_peak, aes(x=sex, y = mean, color = sex), alpha = 0.5, stroke = NA, width = 0.25, height = 0.1) + 
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se),
                width = 0.3,
                color = "black",
                size = 0.5) +  
  theme_prism() + 
  ylab("Peak Behavioral Intoxication Score") +
  scale_x_discrete(labels=c("female" = "Female", 
                            "male" = "Male")) + 
  xlab("Sex") + 
  scale_y_continuous(limits = c(0,5), expand = c(0, 0)) + 
  theme(legend.position = "none")

ggsave(
  "/behav_intox_peak.png",
  path = results_path,
  height = 4,
  width = 2.5
  )

```

**Figure 2A: Ethanol Dose Per Day**

```{r}
#Creating Data Frame 
dose_day <- binge_stats_filtered %>%
  select(6, 13:16, 50) %>% #selecting relevant columns of binge_stats
  pivot_longer(
    cols = -c(unique_id, sex),
    names_to = "time",
    values_to = "dose"
  ) 

#creating summary statistics 
dose_day_grouped <- dose_day %>%
  group_by(sex, time) %>%
  get_summary_stats(dose, show = c("mean", "sd", "se", "ci"))
dose_day_grouped

#creating model and running anova
model_dose <- lmer(dose ~ sex * time + (1 | unique_id), data = dose_day)
model_dose_anova <- anova(model_dose)
print(model_dose_anova)

#posthoc test
emmeans_dose <- emmeans(model_dose, ~sex | time)
emmeans_dose_results <- pairs(emmeans_dose)
print(emmeans_dose_results)

#graph of behavioral intoxication over time 
ggplot(data = dose_day_grouped, aes(x = time, y = mean, color = sex, group = sex)) + 
  geom_line(size = 1) + 
  geom_point() + 
  geom_errorbar(aes(y = mean, ymin = mean - se, ymax = mean + se, fill = sex), alpha = .5, width = 0.25) + 
  ylim(-.25, 15) + 
  scale_x_discrete(labels=c("dose_day1" = "Day 1", 
                            "dose_day2" = "Day 2",
                            "dose_day3" = "Day 3",
                            "dose_day4" = "Day 4"
                            )) +
  ylab("Mean Ethanol Dose (g/kg)") + 
  xlab("Ethanol Treatment") + 
  theme_prism() +
  theme(legend.position = "none")

ggsave(
  "/day_dose.png",
  path = results_path,
  height = 5,
  width = 8.8
  )
```

**Figure 2B: Mean Dose**

```{r}
#calculating summary statistics 
dose_day_grouped_2<- dose_day %>%
  group_by(sex) %>%
  get_summary_stats(dose, show = c("mean", "sd", "se", "ci"))
dose_day_grouped_2

#data frame for the jittered points
dose_day_mean <- dose_day %>%
  group_by(unique_id, sex) %>%
  summarize(mean = mean(dose))

#t.test
dose_t.test <- t.test(dose ~ sex, data = dose_day)
dose_t.test

#mean dosegraph 
ggplot(data = dose_day_grouped_2, aes(x = sex, y = mean)) + 
  geom_bar(aes(fill = sex), stat = "summary", fun = "mean", alpha = 0.5, width = 0.5, color = "black") +
  geom_jitter(data = dose_day_mean, aes(x=sex, y = mean, color = sex), alpha = 0.5, stroke = NA, width = 0.25, height = 0.1) + 
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se),
                width = 0.3,
                color = "black",
                size = 0.5) +   
  theme_prism() + 
  ylab("Mean Dose (g/kg/day") +
  scale_x_discrete(labels=c("female" = "Female", 
                            "male" = "Male")) + 
  xlab("Sex") + 
  scale_y_continuous(limits = c(0,15), expand = c(0, 0)) + 
  theme(legend.position = "none")

ggsave(
  "/dose.png",
  path = results_path,
  height = 4,
  width = 3
  )

```

**Figure 2C: BEC**

```{r}
#Creating Data Frame 
bec <- binge_stats_filtered %>%
  select(6, 19, 50)  #selecting relevant columns of binge_stats
 
#calculating summary statistics 
bec_grouped<- bec %>%
  group_by(sex) %>%
  get_summary_stats(mean_bec, show = c("mean", "sd", "se", "ci"))
bec_grouped

#t.test
dose_t.test <- t.test(mean_bec ~ sex, data = bec)
dose_t.test

#mean bec 
ggplot(data = bec_grouped, aes(x = sex, y = mean)) + 
  geom_bar(aes(fill = sex), stat = "summary", fun = "mean", alpha = 0.5, width = 0.5, color = "black") +
  geom_jitter(data = bec, aes(x=sex, y = mean_bec, color = sex), alpha = 0.5, stroke = NA, width = 0.25, height = 0.1) + 
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se, color = sex),
                width = 0.3,
                color = "black",
                size = 0.5) +   
  theme_prism() + 
  ylab("BEC (mg/dl)") +
  scale_x_discrete(labels=c("female" = "Female", 
                            "male" = "Male")) + 
  xlab("Sex") + 
  scale_y_continuous(limits = c(0,700), expand = c(0, 0)) + 
  theme(legend.position = "none")

ggsave(
  "/bec.png",
  path = results_path,
  height = 4,
  width = 3
  )


```

**Figure 3A: Withdrawal Over Time**

```{r}
#Creating Data Frame 
withdrawal <- binge_stats_filtered %>%
  select(6, 32:48, 50) %>% #selecting relevant columns of binge_stats
  pivot_longer(
    cols = -c(unique_id, sex),
    names_to = "time",
    values_to = "with_score"
  ) 

#creating summary statistics 
withdrawal_grouped <- withdrawal %>%
  group_by(sex, time) %>%
  get_summary_stats(with_score, show = c("mean", "sd", "se", "ci"))
print(withdrawal_grouped, n=100)

#creating model and running anova
model_withdrawal <- lmer(with_score ~ sex * time + (1 | unique_id), data = withdrawal)
model_withdrawal_anova <- anova(model_withdrawal)
print(model_withdrawal_anova)

#posthoc test
emmeans_withdrawal <- emmeans(model_withdrawal, ~sex | time)
emmeans_withdrawal_results <- pairs(emmeans_withdrawal)
print(emmeans_withdrawal_results)

#graph of behavioral intoxication over time 
ggplot(data = withdrawal_grouped, aes(x = time, y = mean, color = sex, group = sex)) + 
  geom_line(size = 1) + 
  geom_point() + 
  geom_errorbar(aes(y = mean, ymin = mean - se, ymax = mean + se, fill = sex), alpha = .5, width = 0.5) + 
  ylim(-.25, 3) + 
  scale_x_discrete(labels=c("with_10" = "10", 
                            "with_11" = "11",
                            "with_12" = "12", 
                            "with_13" = "13",
                            "with_14" = "14", 
                            "with_15" = "15",
                            "with_16" = "16", 
                            "with_17" = "17",
                            "with_18" = "18", 
                            "with_19" = "19",
                            "with_20" = "20", 
                            "with_21" = "21",
                            "with_22" = "22", 
                            "with_23" = "23",
                            "with_24" = "24", 
                            "with_25" = "25",
                            "with_26" = "26"
                            )) +
  ylab("Mean Withdrawal Score") + 
  xlab("Hours Since Last Ethanol Dose") + 
  theme_prism() +
  theme(legend.position = "none")

ggsave(
  "/withdrawal_time.png",
  path = results_path,
  height = 5,
  width = 8.8
  )

```

**FIgure 3B: Mean Withdrawal**

```{r}
#calculating summary statistics 
withdrawal_grouped_average<- withdrawal %>%
  group_by(sex, unique_id) %>%
  summarize(mean_with_score = mean(with_score, na.rm = TRUE)) #calculating max intox score 

#calculating summmary statistics 
withdrawal_grouped_average_stats <- withdrawal_grouped_average %>%
  get_summary_stats(mean_with_score, show = c("mean", "sd", "se", "ci"))
withdrawal_grouped_average_stats

#t.test
withdrawal_average_t.test <- t.test(mean_with_score ~ sex, data = withdrawal_grouped_average)
withdrawal_average_t.test

#mean withdrawal intoxication graph 
ggplot(data = withdrawal_grouped_average_stats, aes(x = sex, y = mean)) + 
  geom_bar(aes(fill = sex), stat = "summary", fun = "mean", alpha = 0.5, width = 0.5, color = "black") + 
  geom_jitter(data = withdrawal_grouped_average, aes(x=sex, y = mean_with_score, color = sex), alpha = 0.5, stroke = NA, width = 0.25) + 
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se),
                width = 0.2,
                color = "black",
                size = 0.5) +  
  theme_prism() + 
  ylab("Mean Withdrawal Score") +
  scale_x_discrete(labels=c("female" = "Female", 
                            "male" = "Male")) + 
  xlab("Sex") + 
  scale_y_continuous(limits = c(0,5), expand = c(0, 0)) + 
  theme(legend.position = "none")

ggsave(
  "/withdrawal_mean.png",
  path = results_path,
  height = 4,
  width = 2.5
  )

```

**Figure : Peak Withdrawal**

```{r}
#getting the max intox score
withdrawal_grouped_peak<- withdrawal %>%
  group_by(sex, unique_id) %>%
  summarize(max_with_score = max(with_score, na.rm = TRUE)) #calculating max intox score 
  
#calculating summmary statistics 
withdrawal_grouped_peak_stats <- withdrawal_grouped_peak %>%
  get_summary_stats(max_with_score, show = c("mean", "sd", "se", "ci"))
withdrawal_grouped_peak_stats

#t.test
withdrawal_peak_t.test <- t.test(max_with_score ~ sex, data = withdrawal_grouped_peak)
withdrawal_peak_t.test

#mean behavioral intoxication graph 
ggplot(data = withdrawal_grouped_peak_stats, aes(x = sex, y = mean)) + 
  geom_bar(aes(fill = sex), stat = "summary", fun = "mean", alpha = 0.5, width = 0.5, color = "black") + 
  geom_jitter(data = withdrawal_grouped_peak, aes(x=sex, y = max_with_score, color = sex), alpha = 0.5, stroke = NA, width = 0.25, height = 0.1) + 
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se),
                width = 0.3,
                color = "black",
                size = 0.5) +  
  theme_prism() + 
  ylab("Peak Withdrawal Score") +
  scale_x_discrete(labels=c("female" = "Female", 
                            "male" = "Male")) + 
  xlab("Sex") + 
  scale_y_continuous(limits = c(0,5), expand = c(0, 0)) + 
  theme(legend.position = "none")

ggsave(
  "/withdrawal_peak.png",
  path = results_path,
  height = 4,
  width = 2.5
  )


# Check for NA values in the with_score column
sum(is.na(withdrawal$with_score))

# Check for NA values in the grouping columns
sum(is.na(withdrawal$sex))
sum(is.na(withdrawal$unique_id))

# Inspect the data to understand the distribution of NA values
summary(withdrawal)


```

**Female Correlations**

```{r}
female_behav_intox_average <- behav_intox_average |> 
  filter(sex == "female") |> 
  arrange(unique_id)

#Average BEC
female_bec <- bec |> 
  filter(sex == "female") |> 
  arrange(unique_id) |>
  as_tibble()

#average WD
female_withdrawal_average <- withdrawal_grouped_average |> 
  filter(sex == "female") |> 
  arrange(unique_id)

#peak WD
female_withdrawal_peak <- withdrawal_grouped_peak |> 
  filter(sex == "female") |> 
  arrange(unique_id)

#combining data so that varaibles match to subject
female_correlation_variables <- female_behav_intox_average |>
  inner_join(female_bec) |>
  inner_join(female_withdrawal_average) |>
  inner_join(female_withdrawal_peak)

#correlation analysis 

female_behav_intox_vs_meanWD <- cor.test(female_correlation_variables$mean, female_correlation_variables$mean_with_score, method = "spearman")

female_behav_intox_vs_peakWD <- cor.test(female_correlation_variables$mean, female_correlation_variables$max_with_score, method = "spearman")

female_BEC_vs_meanWD <- cor.test(female_correlation_variables$mean_bec, female_correlation_variables$mean_with_score, method = "spearman")
  
female_BEC_vs_peakWD <- cor.test(female_correlation_variables$mean_bec, female_correlation_variables$max_with_score,method = "spearman")

female_behav_intox_vs_BEC <- cor.test(female_correlation_variables$mean, female_correlation_variables$mean_bec, method = "spearman")

female_behav_intox_vs_meanWD
female_behav_intox_vs_peakWD
female_BEC_vs_meanWD
female_BEC_vs_peakWD
female_behav_intox_vs_BEC

```

**Male Correlations**

```{r}
male_behav_intox_average <- behav_intox_average |> 
  filter(sex == "male") |> 
  arrange(unique_id)

#Average BEC
male_bec <- bec |> 
  filter(sex == "male") |> 
  arrange(unique_id) |>
  as_tibble()

#average WD
male_withdrawal_average <- withdrawal_grouped_average |> 
  filter(sex == "male") |> 
  arrange(unique_id)

#peak WD
male_withdrawal_peak <- withdrawal_grouped_peak |> 
  filter(sex == "male") |> 
  arrange(unique_id)

#combining data so that varaibles match to subject
male_correlation_variables <- male_behav_intox_average |>
  inner_join(male_bec) |>
  inner_join(male_withdrawal_average) |>
  inner_join(male_withdrawal_peak)

#correlation analysis 

male_behav_intox_vs_meanWD <- cor.test(male_correlation_variables$mean, male_correlation_variables$mean_with_score, method = "spearman")

male_behav_intox_vs_peakWD <- cor.test(male_correlation_variables$mean, male_correlation_variables$max_with_score, method = "spearman")

male_BEC_vs_meanWD <- cor.test(male_correlation_variables$mean_bec, male_correlation_variables$mean_with_score, method = "spearman")

male_BEC_vs_peakWD <- cor.test(male_correlation_variables$mean_bec, male_correlation_variables$max_with_score,method = "spearman")

male_behav_intox_vs_BEC <- cor.test(male_correlation_variables$mean, male_correlation_variables$mean_bec, method = "spearman")

male_behav_intox_vs_meanWD
male_behav_intox_vs_peakWD
male_BEC_vs_meanWD
male_BEC_vs_peakWD
male_behav_intox_vs_BEC
```

Chi-Square Test: Convulsions

```{r}
#creating a new column showing who didn't and did have a convulsions 
withdrawal_grouped_peak_convul <- withdrawal_grouped_peak |> 
  ungroup() |>
  mutate(convulsions = case_when(max_with_score < 3.8 ~ "no",
                                 max_with_score >= 3.8 ~ "yes"))
                                
#table showing the percentage of convulsions 
observed_convulsions <- table(withdrawal_grouped_peak_convul$sex, withdrawal_grouped_peak_convul$convulsions)
observed_convulsions

#chi square test looking at convulsions between males and females
chisq.test(observed_convulsions)

```

**Percentage of Convulsions:**

**Female**

```{r}
female_convlusions <- withdrawal_grouped_peak |> 
  filter(sex == "female") |> 
  arrange(unique_id) |> 
  filter(max_with_score == "3.8") |>
  nrow()
# number of female WD scores 
female_wdn <- withdrawal_grouped_peak |> 
  filter(sex == "female") |> 
  arrange(unique_id) |> 
  nrow()

#female convulsions percentage
female_percent_by_group <- (female_convlusions /  female_wdn) * 100
female_percent_by_group


```

**Male**

```{r}
#number of male convulsions
male_convlusions <- withdrawal_grouped_peak |> 
  filter(sex == "male") |> 
  arrange(unique_id) |> 
  filter(max_with_score == "3.8") |>
  nrow()
# number of male WD scores 
male_wdn <- withdrawal_grouped_peak |> 
  filter(sex == "male") |> 
  arrange(unique_id) |> 
  nrow()
#male convulsions percentage
Male_percent_by_group <- (male_convlusions /  male_wdn) * 100
Male_percent_by_group 
```
